Search-based Synthesis of Network Intrusion Detection Rules

### 2019-09-25 paper/paper.tex
pass

Network Intrusion Detection Systems (\nids{}) are software systems
that monitor the network traffic for malicious behavior and act
accordingly by blocking messages or alerting humans about suspicious
events~\cite{Mitchell:2014:SID:2597757.2542049}. \nids{} are typically
placed behind a firewall, vetting the traffic that the firewall did
not block. Various open-source (\eg{}, Snort~\cite{snort} and
Suricata~\cite{suricata}) and commercial implementations (\eg{},
SolarWinds~\cite{solarwinds} and IBM QRadar~\cite{qradar}) exist
today. These systems are very popular in industry to secure local
computer networks given the amount of potential malicious traffic that
exist on the Internet today.

\sloppy \nids{} are typically categorized in two
groups~\cite{kumar2007survey}: 1)~rule-based and 2)~anomaly-based. A
rule-based \nids{} (\aka\ signature-based NIDS) checks if the network
traffic matches a fixed set of
rules. Figure~\ref{fig:synflood-example} shows an example rule of
Suricata~\cite{suricata}, a popular open-source \nids{}. This rule
prescribes a method to capture a denial-of-service attack to a
server~\cite{Douligeris:2004:DAD:987153.987158}. Rule-based \nids{} focus on known attacks whereas
anomaly-based \nids{} focus on unanticipated attacks that could be
observed through suspicious network
traffic~\cite{kumar2007survey,Mitchell:2014:SID:2597757.2542049,cordy-etal-issta19}.

These intrusion detection mechanisms are complementary. Rule-based
NIDS are restricted to a fixed set of rules and not all rules are
simultaneously enabled (for efficiency). Attackers constantly create
new attacks, and manually writing rules is tedious and
error-prone. However, the semantics of the specification language of
rules is clear making it easy for system administrator to understand
them. As such, rule-based systems are very popular in industry.
Anomaly-based NIDS are not restricted to a fixed set of rules; they
learn common behaviors and dynamically detect uncommon behavior. They
typically use machine learning algorithm to model event
sequences~\cite{7579764}. Unfortunately, anomaly-based NIDS can raise
false alarms as behavior models are fundamentally imprecise. To sum
up, both approaches have positive and negative aspects.

@@

This paper proposes a technique, dubbed \tname{}, to automatically
synthesize rules from positive and negative examples (\ie{}, benign
and malicious traffic). The main scenario of application is one where
an anomaly-based NIDS identifies abnormal traffic and signals that
traffic to \tname{} to craft rules that can be used by rule-based
NIDS. Those rules can be distributed to rule databases. The core
motivation is that 1) attackers are productive in creating new ways to
circumvent existing protections and 2) manual creation of rules is
tedious and time-consuming.  Note that anomaly-based NIDS are evolving
pretty quick with the advances in machine learning, but rule-based
NIDS are still extremely popular. Our goal is to harden rule-based
NIDS by automatically creating rules from suspicious behavior detected
by anomaly-based NIDS.

@@

One important step in search-based optimization is to characterize a
candidate solution to the target problem. Candidate solutions are the
objects that the search optimize towards solutions. In this case, a a
candidate solution is an encoding of a Suricata rule.  \tname\ encodes
the rule options segment of a Suricata rule (see
Section~\ref{sec:example-suricata-rules}) as a vector of Option
objects. Some options are discarded as they only serve for
documentation. For example, the options \CodeIn{msg},
\CodeIn{classtype}, and \CodeIn{misc-activity}, appearing in the
example on Figure~\ref{fig:synflood-example}, have no influence in
triggering the actions associated with rules. The option
\CodeIn{flags}, however, is relevant to trigger the action. \tname{}
models that option with a data structure storing two fields, one of
character type and another of integer type. This enables \tname{} to
mutate each element separately. Similar encodings are used in modeling
the other relevant options. We found what is (ir)relevant by
inspecting Suricata code for the presence of \CodeIn{eval} functions,
which are invoked from a main look to process every Suricata rule. The
presence of such a function for a given option indicates that it is
relevant for triggering some action of some rule.

Options can be specific to a given protocol or general and they can be
influential on triggering an action or
not. Figure~\ref{fig:distribution-rules-protocol} shows the
distributions of these options per protocol and their relevance for
triggering rule actions.

>>>>

Network Intrusion Detection Systems (\nids{}) are software systems
that monitor the network traffic for malicious behavior and act
accordingly by blocking messages or alerting humans about suspicious
events~\cite{Mitchell:2014:SID:2597757.2542049}. \nids{} are typically
placed behind a firewall, vetting the traffic that the firewall did
not block. Various open-source (\eg{}, Snort~\cite{snort} and
Suricata~\cite{suricata}) and commercial implementations (\eg{},
SolarWinds~\cite{solarwinds} and IBM QRadar~\cite{qradar}) exist
today. These systems are very popular in industry to secure local
computer networks given the amount of potential malicious traffic that
exist on the Internet.

\sloppy \nids{} are typically categorized in two
groups~\cite{kumar2007survey}: rule-based and anomaly-based \nids. A
rule-based intrusion detector\footnote{\aka\ signature-based intrusion
  detector.} checks if the network traffic matches a fixed set of
rules. Figure~\ref{fig:synflood-example} shows an example rule of
Suricata~\cite{suricata}, a popular open-source \nids{}. This rule
prescribes a method to capture a denial-of-service
attack~\cite{understanding-dos} to a server by matching specific
conditions about the network traffic. Relevant conditions appear in
bold. Rule-based \nids{} focus on known attacks whereas anomaly-based
\nids{} focus on unknown attacks that could be observed by first
learning the regular traffic and then capturing suspicious
traffic~\cite{kumar2007survey,Mitchell:2014:SID:2597757.2542049,cordy-etal-issta19}.

These two kinds of intrusion detection mechanisms are
complementary. Conceptually, rule-based NIDS could miss attacks (as
rulesets can become outdated) whereas anomaly-based NIDS can raise
false alarms (as the detection algorithms use are approximate).
Rule-based NIDS are restricted to a fixed set of rules and attackers
constantly create new strategies. Furthermore, manually writing rules
to capture those attacks is tedious and error-prone. As consequence,
IT-security companies advertize paid rulesets for consumers to
buy~\cite{proofpoint-etpro,snort-rule-subscriptions}.  Despite this
fundamental issue, rule-based systems are tremendously popular in
industry today~\Fix{cite}. Anomaly-based NIDS, in contrast, are not
restricted to a fixed set of rules. They learn usage behavior from
positive traffic and, based on that, alert uncommon
behavior~\cite{7579764}. Unfortunately, anomaly-based NIDS are
fundamentally imprecise. To sum up, both approaches have positive and
negative aspects.

@@

This paper proposes a technique, dubbed \tname{}, to close the gap
between these two strategies. \tname{} automatically synthesizes rules
from positive and negative examples (\ie{}, benign and malicious
traffic). A scenario of application of \tname{} is one where an
anomaly-based NIDS identifies abnormal traffic and signals that
traffic to \tname{} to craft rules that can be used by rule-based
NIDS. Note that anomaly-based NIDS are evolving pretty quick with the
advances in machine learning, but rule-based NIDS are still extremely
popular.  \tname{} could also leverage existing databases of malicious
traffic to synthesize rules.  Regardless of how the negative traffic
is produced (out of scope of this paper), those rules can be
distributed to rule databases for free. Our goal is to harden the
protection of NIDS by automatically synthesizing rules from negative
examples.

%%  The
%% core motivation is that 1) attackers are productive in creating new
%% ways to circumvent existing protections and 2) manual creation of
%% rules is tedious and time-consuming.

@@

One important step in search-based optimization is to characterize a
candidate solution to the target problem. Candidate solutions are the
objects that the search optimize towards solutions. In this case, a 
candidate solution is an encoding of a Suricata rule.  \tname\ encodes
the rule options segment of a Suricata rule (see
Section~\ref{sec:example-suricata-rules}) as a vector of Option
objects. Some options are discarded as they only serve for
documentation. For example, the options \CodeIn{msg},
\CodeIn{classtype}, and \CodeIn{misc-activity}, appearing in the
example on Figure~\ref{fig:synflood-example}, have no influence in
triggering the actions associated with the rule. The option
\CodeIn{flags}, however, is relevant to trigger the action. \tname{}
models that option with a data structure storing two fields, one of
character type and another of integer type. This enables \tname{} to
mutate each element separately. Similar encodings are used in modeling
the other relevant options. We found what options are (ir)relevant by
inspecting Suricata code for the presence of \CodeIn{eval} functions,
which are invoked from a main loop to process each action rule. The
presence of such a function for a given option indicates that it is
relevant for triggering some action of some rule.

Options can be specific to a given protocol or general and they can
influence or not triggering an
action. Figure~\ref{fig:distribution-rules-protocol} shows the
distributions of these options per protocol and their relevance for
triggering rule actions.
